APOGEE CATALOG OF 10,000+ M DWARFS MEASURED BY DATA-DRIVEN MODELS

### 2018-07-26 paper/draft1/draft1.tex
backup

The Cannon \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (APOGEE; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mu m$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H], and chemical abundances) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars. Using a training set of 51 M dwarfs with spectral types ranging M0-M9 obtained from SDSS optical spectra, we demonstrate that The Cannon can infer spectral types to a precision of 0.6 types. We then use 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 to train a two-parameter model precise to 68K and 0.07 dex respectively. Finally we apply our model to determine spectral subtypes, temperatures and metallicities for 10,000+ M dwarfs in the survey and compare our results to APOGEE pipeline measurements.

@@

\keywords{stars: Mdwarfs -- stellar parameters}

@@

% \begin{enumerate}
Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.

@@

% Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them:

@@

The launch of the (APOGEE; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy.

@@

Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: models dont accurately reproduce data; empirical standards are relative not absolute.

@@

Why The Cannon approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016})

@@

The physical parameter sample consists of 30 sources with reference labels distributed over $3072 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 drawn from \citealt{Mann:2015} (hereafter M15). The M15 catalog in total contains 183 sources selected from CONCH-SHELL \citep{Gaidos:2013} and with infrared spectra taken in follow up from the SpeX Spectograph. Effective temperatures are measured by comparison of BT-SETTL atmospheric models \citep{Allard:2011} to optical spectra. A subsample of 29 sources with measured angular diameters and bolometric fluxes from long-baseline optical interferometry are used to callibrate effective temperature measurements. Relative uncertainty from the callibration sample is measured to be \color{red}{$\sim$30K (check this)}\color{black}, which added in quadrature to the scatter from model comparison gives a total reported uncertainty of $\sim$60K in T$_{\rm eff}$.

@@

We note that surface gravity (logg) and rotational velocity (vsini) parameters are not included as training labels because...

\subsection{Test Sample Selection}

@@

We use the aspcapStar spectra (see \citealt{Perez:2016} for a complete description of the pipeline).

APOGEE data products come in several types of files: ap1D, apVisit, apStar, and aspcapStar. SNR of stars in our sample.

@@

Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector \emph{l} and and coefficient vector \emph{$\theta$}. So for each star \emph{n}, at pixel wavelength \emph{$\lambda$}, we state that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
\begin{equation}
	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + {\rm noise}
\end{equation}

Where the noise model is ${\rm noise}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, which the root mean sum of (1) the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and (2) the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$. Quadratic label vectors:
	\[\ell_{n} = [1, \, SPT, \, SPT^{2}] \]
	\[\ell_{n} = [1, \, T_{\rm eff}, \, [Fe/H], \, T_{\rm eff} \cdot [Fe/H], \, [Fe/H]^{2}] \]

The training step is the stage

@@

For each model, two tests of consistency were used to assess the validity of our derived test labels. The self-test comparing training labels and test labels where precision and bias of the model are calculated as the standard deviation and mean of the difference in training and test labels respectively.

Some note on what lines vary significantly with parameter?

\subsubsection{Spectral Type Model}
We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type. We note however, that the training sample is distributed mostly towards earlier type sources, with only one M8 and one M9 source.

\subsubsection{Temperature/Metallicity Model}
For the physical parameter model we trained The Cannon on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 68K/0.07dex respectively, similar to the precision of the training uncertainties of 60K/0.08dex.

We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), however the model

@@

\subsection{Test Sample}
Some validation of our test label results...

@@

\item[-] The Cannon as a test of evaluating training parameter accuracy? 
\end{enumerate}

@@

\begin{figure}[ht]
\begin{center}
\includegraphics[width=15cm]{figures/self_and_validation_test_spt_snr.png}
\end{center}
\caption{Consistency test for the West-trained spectral type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}.} \label{fig:west_validation}
\end{figure}

\begin{figure}[ht]
\begin{center}
\includegraphics[width=15cm]{figures/self_and_validation_test_teff_snr.png}
\includegraphics[width=15cm]{figures/self_and_validation_test_feh_snr.png}
\end{center}
\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test \textit{(left)}, and leave-one-out cross validation \textit{(right)}. Sources with strong rotational broadening (2M07171706−0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles.} \label{fig:mann_validation}
\end{figure}

@@

%---------------------------
\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/demo_derivatives_teff1.png}
\includegraphics[width=16cm]{figures/demo_derivatives_teff2.png}
\includegraphics[width=16cm]{figures/demo_derivatives_teff3.png}
\end{center}
\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities. \textit{Third panel:} Derivative of the Cannon model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463 K$}. \label{fig:demo_teff}
\end{figure}

@@

%---------------------------
\begin{figure}[ht]
\plotone{figures/cannon_chi_dist.png}
\caption{The distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE-Gaia cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$ \textit{(right)}, and the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}).}
\plotone{figures/safe_selection.png}
\caption{Safe sample selection.}
\end{figure}

>>>>

The Cannon \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (APOGEE; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mu m$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H], and chemical abundances) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the APOGEE pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars. Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49 we train a two-parameter model precise to 68K and 0.07 dex respectively. We then train a one-dimensional spectral type model using 51 M dwarfs ranging M0-M9 obtained from SDSS optical spectra, and demonstrate that The Cannon can infer spectral types to a precision of 0.6 types. Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,084 M dwarfs in the survey and compare our results to APOGEE pipeline measurements, and find our results to be significantly more precise and accurate than ASPCAP, with scatters of $\sim80-100K$ comapred to color-temperature relations from several literature sources.

@@

% \keywords{stars: Mdwarfs -- stellar parameters}

@@

% \begin{enumerate}
Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}.

@@

@@

The launch of the (APOGEE; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy. \color{gcolor}{What about APOGEE's wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?}\color{black}

@@

Why is there a problem labeling M dwarfs (and especially for high res APOGEE spectra)? Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.

@@

Why The Cannon approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}) \color{gcolor}{For Hogg to articulate...}\color{black}

@@

The physical parameter sample consists of 87 sources with reference labels distributed over $2859 < T_{\rm eff} < 4131K$, and -0.48 $<$ [Fe/H] $<$ 0.49, 41 sources of which are drawn from \citealt{Mann:2015} (hereafter M15), and 46 of which are part of an previously unpublished extension sample. The M15 catalog in total contains 183 sources selected from CONCH-SHELL \citep{Gaidos:2013} and with infrared spectra taken in follow up from the SpeX Spectograph. Effective temperatures are measured by comparison of BT-SETTL atmospheric models \citep{Allard:2011} to optical spectra. A subsample of 29 sources with measured angular diameters and bolometric fluxes from long-baseline optical interferometry are used to callibrate effective temperature measurements. Relative uncertainty from the callibration sample is measured to be \color{red}{$\sim$30K (check this)}\color{black}, which added in quadrature to the scatter from model comparison gives a total reported uncertainty of $\sim$60K in T$_{\rm eff}$. Derivation of [Fe/H] and reported uncertainty are...

@@

We note that surface gravity (logg) and rotational velocity (vsini) parameters are not included as training labels because...
\color{gcolor}{MANN: comments to add? particularly about the extension sample?}\color{black}

@@

APOGEE data products: ap1D, apVisit, apStar, and aspcapStar. We use the aspcapStar spectra, which contains the continuum-normalized, rest frame shifted, co-added spectrum of all observed epochs (see \citealt{Perez:2016} for a complete description of the pipeline). SNR of stars in our sample...

@@

Following from the procedure of \citealt{Ness:2015} and \citealt{Ho:2017a} we adopt a linear model that assumes the flux at each pixel of the spectrum can be parameterized as a function of a label vector \emph{l} and and coefficient vector \emph{$\theta$}. So for each star \emph{n}, at pixel wavelength \emph{$\lambda$}, we state that the measured flux for a star at a given pixel is the sum of the coefficient and label product, and observational noise:
\begin{equation}
	f_{n\lambda} = \theta_{\lambda}^{T} \cdot \ell_{n} + {\rm noise}
\end{equation}
We use the noise model is ${\rm noise}=[s_{\lambda}^2 + \sigma_{n\lambda}^2]\xi_{n\lambda}$, where the bracketed term is the root mean sum of the intrinsic scatter of the model at each pixel \emph{$s_{\lambda}$}, and the uncertainty due to instrumental effects \emph{$\sigma_{n\lambda}$}, which is then multiplied times a Gaussian random number $\xi_{n\lambda} \sim \mathcal{N} (0,1)$.

Quadratic label vectors:
	\[\ell_{n} = [1, \, SPT, \, SPT^{2}] \]
	\[\ell_{n} = [1, \, T_{\rm eff}, \, [Fe/H], \, T_{\rm eff} \cdot [Fe/H], \, [Fe/H]^{2}] \]

The training step description...

Test step description...

@@

For each model, two tests of consistency were used to assess the validity of our derived test labels: (1) the self-test comparing training labels and test labels, and (2) the leave-one-out cross validation test, in which we train a model on all sources but \emph{n}, then apply the \emph{N-1} source trained model to get the labels for \emph{n}. Precision (scatter) and bias of the model for each test are calculated as the standard deviation and mean of the difference in training and test labels respectively.

Another mode of analysis we can look at with The Cannon is how the derivative of the model changes with respect to given training parameter, which makes our model interpretable for discovering or verifying atomic or molecular lines with strong dependence on different physical parameters. As discussed further in sections \ref{subsec:mann_results} and \ref{subsec:west_results}, we show the derivative plots of each of our two models and compute the error of the derivative at each pixel using a jackknife statistic:
	\[\sigma_{\theta,m}^2 = \frac{N-1}{N} \sum^N_{n=1} (\theta_{/n}^m - \theta^m)^2 \]
where $\sigma_{\theta,m}$ is the error at pixel \emph{m}, \emph{N} is the total number of stars in the sample, indexed by n, $\theta$ is the coeffient vector trainined on all N sources, and $\theta_{/n}$ is the coefficient vector trained on \emph{N-1} sources excluding star \emph{n}.

\subsubsection{Temperature/Metallicity Model \label{subsec:mann_results}}
For the physical parameter model we trained The Cannon on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 68K/0.07dex respectively, similar to the original training sample uncertainties of 60K/0.08dex. We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure \ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$Gyr) and likely slowly rotating sources. While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources does not significantly

Some note on what lines vary significantly with parameter...

@@

\subsubsection{Spectral Type Model \label{subsec:west_results}}
We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.6 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type. We note however, that the training sample is distributed heavily towards earlier type sources, with only one M8 and one M9 source.

\subsection{Test Sample}
Out of the total APOGEE DR14 catalog of 277,371 sources, we selected 254,478 sources which were in the cross match of Gaia DR2 (\citealt{Brown:2018}) and applied Gaia color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources. From there we applied the following selection procedure to identify a "safe" sample of M dwarfs:

@@

\item[-] The Cannon as a test of evaluating training parameter accuracy? 
\item[-] Discussion of validation tests and ground truth
\end{enumerate}

@@

\begin{figure}[ht]
\begin{center}
\plotone{figures/self_and_validation_test_spt_snr.png}
\end{center}
\caption{Consistency test for the West-trained spectral type model; self-test (left), and leave-one-out cross validation (right).} \label{fig:west_validation}
\end{figure}

\begin{figure}[ht]
\plotone{figures/self_and_validation_test_teff_snr.png}
\plotone{figures/self_and_validation_test_feh_snr.png}
\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (left), and leave-one-out cross validation (right). Sources with strong rotational broadening (2M07171706−0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles.} \label{fig:mann_validation}
\end{figure}

@@

%---------------------------
\begin{figure}[ht]
\begin{center}
\includegraphics[width=16cm]{figures/demo_derivatives_teff1.png}
\includegraphics[width=16cm]{figures/demo_derivatives_teff2.png}
\includegraphics[width=16cm]{figures/demo_derivatives_teff3.png}
\end{center}
\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities. \textit{Third panel of each plot:} Derivative of the Cannon model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463 K$}. \label{fig:demo_teff}
\end{figure}

@@

%---------------------------
\begin{figure}[ht]
\plotone{figures/cannon_chi_dist.png}
\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources (\citealt{ElBadry:2018}; \citealt{Skinner:2018}); \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE-Gaia cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\pi>0$. We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as "safe". \label{fig:chi_dist}}
\end{figure}

\begin{figure}[ht]
\plotone{figures/cmd_selection.png}
\caption{Color-magnitude cuts.}
\end{figure}

\begin{figure}[ht]
\plotone{figures/safe_selection.png}
\caption{Safe sample selection.}
\end{figure}
